---
title: "551 Case Study Group5"
author: "Ethan Straub, Leila Naderi, Benjamin Fraizer"
date: "2024-09-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

In professional sports, understanding the factors that contribute to success is essential for athletes, coaches, and analysts. In the context of professional golf, performance metrics such as driving distance, putting efficiency, and accuracy in reaching greens are critical in determining a player's earnings. This study focuses on golfers in the Ladies Professional Golf Association (LPGA), with the goal of constructing a model to predict the prize money earned per round (przrnd). By analyzing several key performance metrics, this study aims to identify the predictors that have the greatest influence on a golfer's earnings.

The response variable in this analysis is the prize money per round (przrnd), and several performance and experience-related predictors are considered. In this case study, all predictor variables are quantitative. The variables include the number of rounds played (rounds), average drive distance (avedist), percentage of fairways hit (pctfrwy), percentage of greens in regulation (pctgrn), average number of putts per round (aveputt), average sand shots per round (avesand), and sand save percentage (pctsndsv). The golfer's name (Golfer) is a categorical variable, but it is not used in the analysis as a predictor since it doesn’t affect the outcome. 

The primary objective of this study is to develop a statistical model that explains the relationship between these performance variables and prize money earned. By identifying the most significant predictors, we can provide insights into which aspects of a golfer's game contribute most to their financial success. This information can be invaluable for golfers and their coaches in making data-driven decisions about training and strategy, allowing them to focus on areas that yield the greatest return on investment. Furthermore, this model may help golf analysts and enthusiasts better understand the factors that distinguish top performers in the LPGA.


# Summary Statistics and Graphs

```{r}
library(ggcorrplot)
library(ggplot2) 
library(scales) 
library(ggrepel) 
df = read.csv("LPGA.csv")
# Taking names out of the design matrix
data = df[,-1]
pairs(data)
```

```{r}
corMat = cor(data)
ggcorrplot(corMat)
```

From the pairs plot, we can see that the distribution of the response variable (Prize money per round) is highly skewed right. It may be a good idea to log transform this variable in a model.

There are no exceptionally strong relationships between variables, but there are a few noticeable trends. 

Golfers who hit longer drives on average tend to hit the fairway less. This reflects the trade off between power and accuracy. It is easy to hit a ball far, and it is easy to hit a ball short with a good line. But it is difficult to hit it far and with the right line. We can create a variable that reflects a golfer's long game by multiplying these variables together after standardizing (to make the contribution from each variable the same).

Average sand shots and percent green have a decently large negative correlation. Most sand traps in golf are close to the green. So if a golfer misses a lot of greens, it is likely that they hit sand traps instead.

Here are some other relationships in the data that make intuitive sense. A better golfer who makes more money per round will generally have longer drives, more accurate fairway and green shots, less putts, hit less sand traps, and hit good shots from the sand. 

Surprisingly, the highest correlation between two variables in the dataset is between number of rounds played and prize money per round. Perhaps the better golfers who make more money make it into more tournaments.

The rounds variable can be used to make new variables that are scaled measures of recognizable measures.

Scaled total distance = average distance * number of rounds

Scaled total putts = average putts * number of rounds

Scaled total sand shots = average sand shots * number of rounds

Scaled total fairway shots = percent of fairway shots * number of rounds

Scaled total green shots = percent of green shots * number of rounds

Scaled total sand saves = percent of sand saves * number of rounds

```{r}
data$totDist = data$avedist * data$rounds
data$totPutts = data$aveputt * data$rounds
data$totSand = data$avesand * data$rounds
data$totFrwy = data$pctfrwy * data$rounds
data$totGrn = data$pctgrn * data$rounds
data$totSsv = data$pctsndsv * data$rounds
przrnd = data$przrnd
dta_temp = as.data.frame(scale(data[,c(1:7, 9:14)]))
dta_temp$longGame = dta_temp$avedist * dta_temp$pctfrwy
std_designMat = as.data.frame(scale(dta_temp))

corMat2 = cor(std_designMat)
ggcorrplot(corMat2)
```


# Analysis

```{r}
# Functions

suppressMessages(library(rstanarm))
suppressMessages(library(bayesplot))
suppressMessages(library(loo))

fit = function(Y = przrnd, logY = TRUE, X = std_designMat, prior = "normal", scale = 2.5, p0 = 4){
  fullMat = cbind(X, Y)
  if(prior == "normal"){
    if(logY == TRUE){
    fit = stan_glm(log(Y) ~ . , data = fullMat, 
                 prior=normal(scale=scale), refresh = 0)
    }
    else{
    fit = stan_glm(Y ~ . , data = fullMat, 
                prior=normal(scale=scale), refresh = 0)
    }
  }
  else if(prior == "horseshoe"){
    p = ncol(X)
    n = nrow(X)
    global_scale <- (p0/(p - p0))/sqrt(n)
    if(logY == TRUE){
    slab_scale <- sqrt(0.3/p0)*sd(log(fullMat$Y))
    fit <- stan_glm(log(Y) ~ ., data=fullMat, refresh=0,
    prior=hs(global_scale=global_scale,
    slab_scale=slab_scale))
    }
    else{
    slab_scale <- sqrt(0.3/p0)*sd(fullMat$Y)
    fit <- stan_glm(Y ~ ., data=fullMat, refresh=0,
    prior=hs(global_scale=global_scale,
    slab_scale=slab_scale))
    }
  }
  else{
    print("spelling error on the prior")
  }
  return(fit)
}

plots = function(logY = TRUE, X = std_designMat, coefDist = TRUE, residFit = TRUE, Y = przrnd, prior = "normal", scale = 2.5, p0 = 4 ){
  fit = fit(Y, logY, X, prior, scale, p0)
  if(residFit == TRUE){
    plot(fit$fitted.values, fit$residuals, xlab = "Fitted Values", 
         ylab = "Residuals", abline(h=c(-sigma(fit), 0, sigma(fit)), lwd=3,
                                    lty=c(2, 1, 2), col = "gray"), pch = 16)
  }
  if(coefDist == TRUE){
    pt <- mcmc_areas(as.matrix(fit), pars=vars(-'(Intercept)',-sigma),
                 prob_outer=0.95, area_method = "scaled height") +
    xlim(c(-1.2,1.2))
    pt <- pt + scale_y_discrete(limits = rev(levels(pt$data$parameter))) +        ggtitle("Posterior Distribution of Slopes")
    pt
  }
}


compare = function(Y = przrnd, logY = c(TRUE, TRUE), X = list(std_designMat, std_designMat[,c(2:8,14)]), prior = c("normal", "horseshoe"), scale = c(.25, 1), p0 = c(4,4), kf = FALSE, k = 5){
  numModels = length(prior)
  fit_list = vector("list", length = numModels)
  for(i in 1:numModels){
    fit = fit(Y, logY[i], X[[i]], prior[i], scale[i], p0[i])
    fit_list[[i]] = fit
  }
  objList = vector("list", length = numModels)
  if(kf == TRUE){
    for(i in 1:numModels){
      if(i == 1){
        model_1 = fit_list[[i]]
        objList[[i]] = rstanarm::kfold(model_1, K = k, scale = NULL)
      }
      if(i == 2){
        model_2 = fit_list[[i]]
        objList[[i]] = rstanarm::kfold(model_2, K = k, scale = NULL)
      }
      if(i == 3){
        model_3 = fit_list[[i]]
        objList[[i]] = rstanarm::kfold(model_3, K = k, scale = NULL)
      }
    }
  }
  else{
    for(i in 1:numModels){
      if(i == 1){
        model_1 = fit_list[[i]]
        objList[[i]] = loo(model_1)
      }
      if(i == 2){
        model_2 = fit_list[[i]]
        objList[[i]] = loo(model_2)
      }
      if(i == 3){
        model_3 = fit_list[[i]]
        objList[[i]] = loo(model_3)
      }
    }
  }
  do.call(rstanarm::loo_compare, objList)
}
```


```{r}
plots(X = std_designMat[,c(2:8,14)])

#suppressMessages(plots(X = std_designMat[,c(2:8,14)]))

compare(X = list(std_designMat[,c(2,14)], std_designMat[,c(4, 9, 14)]), p0 = c(2,2))


```

```{r}
Y = as.numeric(przrnd)
df = cbind(std_designMat, Y)
f = stan_glm(log(Y) ~., data = df, refresh = 0)
fl = fit()
#folds <-kfold_split_random(K=4, N=100)
t = loo::kfold(f, K = 2, scale = NULL)
```




# Results and Conclusion


# Appendix: All code for this report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
